---
title: "UFO sightings part 2"
author: "Saurabh Gupta"
date: "14 August 2018"
output:
  pdf_document: default
  word_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width= 8)
```

# Time Series Analysis and Forecast

```{r}
# load time series object created in previous section
load("monthly_ts.rda")
```

```{r}
plot(monthly_ts, main = "Global UFO sightings" ,
     ylab = "Monthly Sightings")
```

Need to subset the data between 1990 and 2000 to pick up the change in trend.

Also split last 12 months into test data. Rest is training.

```{r}
plot(monthly_ts[time(monthly_ts) > 1990.000], main = "Global UFO sightings" ,
     ylab = "Monthly Sightings", type = "l")
```

The trend is stronger after 1995 but data doesn't start from 0.


```{r}
plot(monthly_ts[time(monthly_ts) > 1995.000], main = "Global UFO sightings" ,
     ylab = "Monthly Sightings", type = "l")
```

Use data from 2000

Check acf and pacf


```{r}
# Train data

traindata <- window(monthly_ts, start = c(2000, 1), end = c(2013, 1))

plot(traindata)
```


```{r}
# Test data

testdata <- window(monthly_ts, start = c(2013, 1))

plot(testdata)
```

AR process with Seasonality at lag 12 months.

```{r}
acf(traindata)
```

AR with lags 8 and 10

```{r}
pacf(traindata)
```


check after differencing at lag 12

Trend is diminished before 2010. There is still some autocorrelation.


```{r}

diff_12 <- diff(traindata, lag = 12)

plot(diff_12)
acf(diff_12)
pacf(diff_12)

```

Difference at lag 1

Series is much more stationary but variance is increasing. Try log transformation.

```{r}
diff_1_12 <- diff(diff_12, lag = 1)

plot(diff_1_12)
acf(diff_1_12)
pacf(diff_1_12)


```

Log transformation. Then differencing.

MA(1) and Seasonal MA(1) seems to be present.

```{r}

log_train <- log(traindata)

diff_log_1_12 <- diff(diff(log_train, lag = 12), lag = 1)

plot(diff_log_1_12)
acf(diff_log_1_12)
pacf(diff_log_1_12)

```


Try arima (1, 1, 1) with S = 12

```{r}
mod_111_12 <- arima(traindata, order = c(1, 1, 1), seasonal = list(order = c(1, 1, 1), period = 12), include.mean = F)

mod_111_12
```

Try on log data

AIC is much better on log data.

```{r}
log_mod_111_12 <- arima(log_train, order = c(1, 1, 1), seasonal = list(order = c(1, 1, 1), period = 12), include.mean = F)

log_mod_111_12
```

check residuals

```{r}
acf(log_mod_111_12$residuals)

pacf(log_mod_111_12$residuals)
```


Remove AR1 term for alternative model.

AIC was better with (1,1,1) but se too high for AR1.

```{r}
log_mod_011_12 <- arima(log_train, order = c(0, 1, 1), seasonal = list(order = c(0, 1, 1), period = 12))

log_mod_011_12
```


Check residuals.

Looks reasonable.

```{r}

acf(log_mod_011_12$residuals)

pacf(log_mod_011_12$residuals)

```

Test Predictions

```{r}

testfit <- predict(log_mod_011_12, 12)

testpred <- exp(testfit$pred)

 testCI_L <- exp(testfit$pred - testfit$se*1.96)

 testCI_U <- exp(testfit$pred + testfit$se*1.96)

ts.plot(testdata, testpred, testCI_L, testCI_U,  col = c( "black", "green", "gray", "gray"), lty = c(1,1,2,2))

```

R square

```{r}
# Test R square

RSS <- sum((testpred - testdata[2:13])^2)

TSS <- var(testdata[2:13])*11

((TSS-RSS)/TSS)

```

Test the other model

```{r}

testfit2 <- predict(log_mod_111_12, 12)

testpred2 <- exp(testfit2$pred)

 testCI_L2 <- exp(testfit2$pred - testfit2$se*1.96)

 testCI_U2 <- exp(testfit2$pred + testfit2$se*1.96)

ts.plot(testdata, testpred2, testCI_L2, testCI_U2,  col = c( "black", "green", "gray", "gray"), lty = c(1,1,2,2))

```

R2 is much higher with ARIMA 111 model. I'll use this one for the final model.

```{r}
# Test R square

RSS2 <- sum((testpred2 - testdata[2:13])^2)

((TSS-RSS2)/TSS)


```

Create final model and forecast using ARIMA (1,1,1)(1,1,1)12

Use data from 2000 as trend is stronger.

```{r}

# Use all the data from 2000 Jan
log_ts <- log(window(monthly_ts, start = c(2000, 1)))

ts_model <- arima(log_ts, order = c(1, 1, 1), seasonal = list(order = c(1, 1, 1), period = 12))

ts_model


```

Forecast


```{r}

fit <- predict(ts_model, 12)

pred <- exp(fit$pred)

CI_L <- exp(fit$pred - fit$se*1.96)

CI_U <- exp(fit$pred + fit$se*1.96)

# data since 2000

plotdata <- window(monthly_ts, start = c(2000, 1))

ts.plot(plotdata, pred, CI_L, CI_U,  col = c( "black", "green", "gray", "gray"), lty = c(1,1,2,2),
        main = "Global UFO sightings - 12 month forecast", ylab = "Number of Sightings",
        xlab = "Time (Months)", sub = "Forecast Period: Jun 2014 to May 2015")

legend("topleft", legend = c("Observed Data", "Forecast", "95% Confidence Interval"),
       cex = 0.7, col = c( "black", "green", "gray", "gray"), lty = c(1,1,2,2))

```

Check with forecast package

```{r }

library(forecast)

#ts2000 <- window(monthly_ts, start = c(2000, 1))

ts_model2 <- Arima(log_ts, order = c(1, 1, 1), seasonal = list(order = c(1, 1, 1), period = 12))

ts_model2

```


```{r}
plot(forecast(ts_model2, h =12))
```

